---
title: "PCA Exercises"
author: "Part 1 - Descriptive statistics and correlation analysis"
output:
  pdf_document: default
  html_document: default
---

<br>
<br>

#### Exercises

> 1.
Prove that $cov(x,x) = var(x)$

> 2.
Prove that $cor(x,x) = 1$

> 3.
In a finite sample, what is the difference between: the average difference between an individual and the mean VS the standard deviation (yes it's not the same)?

> 4. 
Compute the Pearson and Spearman correlation coefficient on the different variables from the iris dataset. Is there any differences? Why?

> 5.
With R, create two samples: one where the Pearson coefficient correlation is equal to 1 (but with 2 distinct variables) and one where it's between [-0.1 ; 0.1]. Use a sample size >100.

> 6.
What is the condition on the type of the variables used in order to compute the Pearson (or Spearman) coefficient correlation (it's implicitly said in the course)?

> 7.
What are the condition(s) that make possible that `cov(x,y)` is equal to `cor(x,y)`?

> 8.
Why the Pearson correlation coefficient isn't enough to say that there is a causal link between the variables? 

> 9.
What happens when doing `cor(X)`, where $X$ is a matrix of quantitative variables?

> 10.
*With your own words*, make a short description of what correlation is. 

> 11.
Based on the sample below, explain *with your own words* why 2 variables created the same way are not correlated.

```{r}
x = rnorm(n=1000)
y = rnorm(n=1000)
cor.test(x,y)
```

> 12.
Create a scatter plot with the sample size in the x-axis and the Pearson correlation coefficient in the y-axis for a range of 10 to 500 for the sample size. For this, you have to use the `generate()` function below. Do the same thing with the p-value in the y-axis. What do you see? What are the implications? 

```{r}
generate = function(sample_size){
  
  #generate sample
  x = rnorm(n=sample_size, sd=15)
  y = 1.001*x + rnorm(n=sample_size, sd=30)
  
  #compute correlation for the sample
  correlation = cor.test(x,y)
  p = correlation$p.value
  r = correlation$estimate
  
  return(r) #change it to p for the second plot
}
```
